TBD

### 2011-03-01 papers/ICFP2011/paper.tex
* Quelques notes sur le papier
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@291 403e655e-980a-47bd-8c60-1ee5348bb6b7

Programs are nowadays faster to write and safer to execute, thanks to
high-level programming concepts and expressive type systems. Still,
one aspect of programming didn't change very much---if not at
all---since the early days of high-level programming: we still write
programs in text files and when we are done, we let the compiler check
and compile them.  Though, every programmer knows that writing a
program longer than a few dozen of lines is a highly non-linear task
involving constant experiments, fixes, rollback on previous
modifications etc., all of those eventually validated by some a
posteriori criterion: execution and ``empirical'' test, and
increasingly type-checking. In a nutshell, programmers spend more time
\emph{editing} than \emph{writing}. Yet, this aspect of their
day-to-day workflow is not, or poorly taken into account in a
developer's toolchain:
\begin{itemize}
\item To avoid recompiling a whole program upon a change, we split
  them into files, and (re-)compile them separatly. Very often yet,
  this separation serve different purpose: abstraction, splitting into
  logical units, namespace management\ldots\ Why should these purposes
  coincide?
\item To keep track of past versions of a program, we use text-based
  version control systems that have no awareness of the actual content
  of the file. In a sense, the non-linearity of the task is managed
  outside the scope of the language, by manual, textual
  transformations. Isn't there a lot to gain in making these tools
  syntax-aware, even semantics-aware?
\item As a result, all the semantic information that could be
  available to the programmer and act as a \emph{guidance tool} to
  write programs (e.g. type informations) are only used by the
  type-checker as a \emph{repressive} measure against programming
  errors (type errors).
\end{itemize}

These observations apply to the usual, wide-spread programming
languages, but even if the current toolchain in use up to now was a
reasonable approximation to the incremental interaction with the
compiler intended by the user, they become \emph{a fortiori} necessary
when it come to the new, demanding languages embedding formal
verification aspects, like proof assistants or rich specification
languages. In these emerging languages, interaction with a type- or
proof-checker is unavoidable due to the difficulty of writing correct
proofs without guidance from the system, and the constant modification
of the source makes necessary to observe incrementally the effect of a
small change on the whole edifice. Moreover, as the complexity of the
object to check increases, we cannot afford anymore to rely on the
usual unit of compilation, the file, a need a finer-grained tool.

@@

% The problem (1 page)
% My idea (2 pages)
% The details (5 pages)
% related works (1-2 pages)
%conclusion and further work (0.5 page)

\bibliographystyle{abbrvnat}

>>>>

Programs are nowadays faster to write and safer to execute, thanks to
high-level programming concepts and expressive type systems. Still,
one aspect of programming didn't change very much---if not at
all---since the early days of high-level programming: we still write
programs in text files and when we are done, we let the compiler check
and compile them.  Though, every programmer knows that writing a
program longer than a few dozen of lines is a highly non-linear task
involving constant experiments, fixes, rollback on previous
modifications etc., all of those eventually validated by some a
posteriori criterion: execution and ``empirical'' test, and
increasingly complex type-checking and type inference. In a nutshell,
programmers spend more time \emph{editing} than \emph{writing}. Yet,
this aspect of their day-to-day workflow is not, or poorly taken into
account in a developer's toolchain:
\begin{itemize}
\item To avoid recompiling a whole program upon a change, we split
  them into files, and (re-)compile them separatly. Very often yet,
  this separation serve different purpose: abstraction, splitting into
  logical units, namespace management\ldots\ Why should these purposes
  coincide?

% On propose quelque chose d'encore plus extreme: faire descendre
% les VCS dans le type-checker. Voir deux paras plus loin.
\item To keep track of past versions of a program, we use text-based
  version control systems that have no awareness of the actual content
  of the file. In a sense, the non-linearity of the task is managed
  outside the scope of the language, by manual, textual
  transformations. Isn't there a lot to gain in making these tools
  syntax-aware, even semantics-aware?

% Je comprends mais je ne sais pas trop ou tu veux en venir.
\item As a result, all the semantic information that could be
  available to the programmer and act as a \emph{guidance tool} to
  write programs (e.g. type informations) are only used by the
  type-checker as a \emph{repressive} measure against programming
  errors (type errors).
\end{itemize}

These observations apply to the usual, wide-spread programming
languages, but even if the current toolchain in use up to now was a
reasonable approximation to the incremental interaction with the
compiler intended by the user, they become \emph{a fortiori} necessary
when it come to the new, demanding languages embedding formal
verification aspects, like proof assistants or programming languages
with rich type system. In these emerging languages, interaction with a
type- or proof-checker is unavoidable due to the difficulty of writing
correct proofs without guidance from the system, and the constant
modification of the source makes necessary to observe incrementally
the effect of a small change on the whole edifice. Moreover, as the
complexity of the object to check increases, we cannot afford anymore
to rely on the usual unit of compilation, the file, a need a
finer-grained tool. 
% On pourrait parler de l'utilisation de l'incrementalite pour le type-checking
% parallele et aussi pour le developpement concurrent.

Current text-based version control technology is designed to help
concurrent and collaborative development but are not used inside
compilers or proof checkers. There is a good reason for that: a
representation of changes at the textual level is not structured
enough to be used easily and efficiently by such symbolic tools. They
need instead a more structured representation of changes with a clear
meaning with respect to the language semantics. For instance, if a
formal argument of a function has been renamed in a source file
edition, a textual representation would refer to a change in every
line that contains an occurrence of that formal argument. A proof
kernel would prefer a higher level description of that edition from
which it would be simple to deduce that no proof has to be rechecked
because the old and the new version of the development are
$\alpha$-equivalent.

% Prochain para: LF. Finir sur les petits problemes de LF.

% Expliciter clairement les contributions du papier:
% Une extension de LF avec des variables meta + lambda-selecteur-de-boite. (SLF)
% Un langage pour representer efficacement un ensemble de termes validés. (NLF)
% Une compilation de SLF vers NLF qui vérifie le bon typage.
% Un typage incrémental par proof-search dans NLF.
% La méta-théorie de NLF qui pose des bases pour la formalisation des VCS sémantiques.
% Un prototype.

@@

% The problem (1 page) et illustration sur un exemple utilisant SLF.

% My idea (2 pages) - un peu vague.

% Presentation de NLF.

% SLF vers NLF

% Metathéorie de NLF

% Applications (typage incrémental, quelques premières pistes pour commit/merge)

% Related works (1-2 pages) (à ne surtout pas négliger!)

% Conclusion and further work (0.5 page)
% Dans le future work, en vrac:
%    - un meta-langage au dessus de NLF = un langage de tactique typé pour Coq?
%    - des algorithmes de parsing incrémentaux
%    - modélisation des conflits et de leur résolution dirigée par des types.
%    - intégration dans Coq pour la parallélisation du noyau. 
%   -  certification du checker pour être intégré dans le noyau de Coq 
\bibliographystyle{abbrvnat}
